---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Chen, Grace"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv(file.path("/Users/gracechen/Desktop/401/5", "abalones.csv"), sep=",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS,mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: There is a mix of 2 factor, 7 numeric, and 1 integer variable types within this dataset. The summary output displayed distributional implications for multiple variables. For instance, WHOLE, SHUCK, RINGS, and VOLUME variables has a maximum value drastically higher than the mean value. This may be a cause for potential skewness to the right and outliers present. ***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
table <- table(mydata$SEX, mydata$CLASS)
addmargins(table(mydata$SEX, mydata$CLASS))
barplot(table, legend = rownames(table), beside = TRUE, col = c("red", "blue", "green"), main = "CLASS membership, SEX-differentiated", xlab = "CLASS", ylab="Frequency")
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The sex distribution of abalones seem to be left skewed for males and females while it is right skewed for infants with infants being prominently in CLASS A1 and A2. This makes sense since CLASS is categorized by age with A1 being the youngest. Infants being present in CLASS A4 and A5 stands out to me. Males and females distribution are relatively similar in distribution shape but males have a slightly higher frequency in each bucket. ***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata),200),]
plot(work[, 2:6])
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$VOLUME, mydata$WHOLE, col = mydata$CLASS, xlab="Volume", ylab="Whole Weight", main = "Whole Weight, in grams | VOLUME, cm^3", cex = 0.8, pch=19)
legend(x = "topleft", legend = levels(mydata$CLASS), col = unique(sort(mydata$CLASS)), pch=19)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK, col = mydata$CLASS, xlab="Whole Weight", ylab="Shuck Weight", main = "Shuck Weight, in grams | Whole Weight, in grams", cex = 0.8, pch=19)
legend(x = "topleft", legend = levels(mydata$CLASS), col = unique(sort(mydata$CLASS)), pch=19)
abline(a=0,b=max(mydata$SHUCK/mydata$WHOLE))
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: Both displays show a positive correlation and linear relationship between the two variables. Although the displays are using different variables, in both graphs as the x variable increases, the y variable increases as well. There is more variability in the first graph compared to the second graph whereas the second graph depicts more distinction amongst the different classes. For instance, Class A5 seems to be scattered arbitrarily in the first graph while in the second graph it is more clustered at the bottom. Therefore, it can be concluded that there is more shuck in younger abalones compared to older abalones that have a higher whole weight.***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX == "I"], main='Infant Ratio', col='red', xlab = 'Infant Ratio')
hist(mydata$RATIO[mydata$SEX == "F"], main='Female Ratio', col='blue', xlab = 'Female Ratio')
hist(mydata$RATIO[mydata$SEX == "M"], main='Male Ratio', col='green', xlab = 'Male Ratio')

boxplot(mydata$RATIO[mydata$SEX == "I"], main = 'Infant Ratio', horizontal = FALSE, col='red', xlab = 'Infant Ratio')
boxplot(mydata$RATIO[mydata$SEX == "F"], main = 'Female Ratio', horizontal = FALSE, col='blue', xlab = 'Female Ratio')
boxplot(mydata$RATIO[mydata$SEX == "M"], main = 'Male Ratio', horizontal = FALSE, col='green', xlab = 'Male Ratio')

qqnorm(mydata$RATIO[mydata$SEX == "I"], main = 'Infant Ratio', col = 'red', cex = 0.8)
qqline(mydata$RATIO[mydata$SEX == "I"])
qqnorm(mydata$RATIO[mydata$SEX == "F"], main = 'Female Ratio', col = 'blue', cex = 0.8)
qqline(mydata$RATIO[mydata$SEX == "F"])
qqnorm(mydata$RATIO[mydata$SEX == "M"], main = 'Male Ratio', col = 'green', cex = 0.8)
qqline(mydata$RATIO[mydata$SEX == "M"])
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: As shown the QQ plots, all three sex classes seem to be following a relatively normal distribution. However, there are also outliers present for all three sexes as depicted in the top right of the QQ plots that do not follow along the qqline. This is further supported by the histograms with all three of the distributions skewed right as well as the outliers shown in the boxplots. ***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
infant_ratio <- boxplot.stats(x = mydata[mydata$SEX == "I", "RATIO"])$out
infant_extreme_ratio <- boxplot.stats(x = mydata[mydata$SEX == "I", "RATIO"], coef = 3.0)
infant_ratio_df <- mydata[mydata$SEX == "I" & mydata$RATIO %in% infant_ratio, ]
female_ratio <- boxplot.stats(x = mydata[mydata$SEX == "F", "RATIO"])$out
female_extreme_ratio <- boxplot.stats(x = mydata[mydata$SEX == "F", "RATIO"], coef = 3.0)
female_ratio_df <- mydata[mydata$SEX == "F" & mydata$RATIO %in% female_ratio, ]
male_ratio <- boxplot.stats(x = mydata[mydata$SEX == "M", "RATIO"])$out
male_extreme_ratio <- boxplot.stats(x = mydata[mydata$SEX == "M", "RATIO"], coef = 3.0)
male_ratio_df <- mydata[mydata$SEX == "M" & mydata$RATIO %in% male_ratio, ]
all_df <- rbind(infant_ratio_df, female_ratio_df, male_ratio_df)
kable(all_df)
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There are 8 outliers present in infants, 5 outliers present in females, and 6 outliers present in males. While most of these outliers are considered mild, there is one outlier in both infants and females that would considered extreme outliers. Most of the outliers can be attributed to infant outliers or young abalones in early age classifications.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow = c(2, 2))
boxplot(mydata$VOLUME ~ mydata$CLASS, main = 'VOLUME Differentiated by CLASS', horizontal = FALSE, col='red', xlab = 'CLASS', ylab= 'VOLUME (cm^3)')
boxplot(mydata$WHOLE ~ mydata$CLASS, main = 'WHOLE Differentiated by CLASS', horizontal = FALSE, col='blue', xlab = 'CLASS', ylab= 'WHOLE (g)')
plot(mydata$RINGS, mydata$VOLUME, xlab="RINGS", ylab="VOLUME (cm^3)", main = "VOLUME vs RINGS", col = 'red', cex = 0.8)
plot(mydata$RINGS, mydata$WHOLE, xlab="RINGS", ylab="WHOLE (g)", main = "WHOLE vs RINGS", col = 'blue', cex = 0.8)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: The variables VOLUME and WHOLE would be poor predictors of age. As shown in the boxplots, classes A3, A4, and A5 are all concentrated around a similar range for both volume and whole. The whiskers for each class are also very long and overlap with the other classes, making it very difficult to differentiate between the three classes. Moreover, this trend continues on the scatterplots. The right side of graph shows little to no correlation between the number of rings and both the volume and whole. Again, this makes it challenging to make distinctions between the rings. However, it looks like this may work as decent predictors of age for younger abalones. In the boxplots, the boxes and whiskers are smaller, making the distinction between classes A1 and A2 evident for both volume and whole. In the scatterplot, there also exists a more of an upward trend in the left side of a graph.  ***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
volume <- aggregate(VOLUME~SEX+CLASS,data = mydata, mean)
shuck <- aggregate(SHUCK~SEX+CLASS,data = mydata, mean)
ratio <- aggregate(RATIO~SEX+CLASS,data = mydata, mean)
matrix(volume$VOLUME, 3, 5, dimnames=list(c('Infant','Female','Male'), c('A1','A2','A3','A4','A5')))
matrix(shuck$SHUCK, 3, 5, dimnames=list(c('Infant','Female','Male'), c('A1','A2','A3','A4','A5')))
matrix(ratio$RATIO, 3, 5, dimnames=list(c('Infant','Female','Male'), c('A1','A2','A3','A4','A5')))

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
interaction.plot(mydata$CLASS, mydata$SEX, mydata$RATIO, fun=mean, col = c('red', 'blue', 'green'), main='Mean Ratio per Class', xlab='CLASS', ylab='MEAN RATIO', trace.label='SEX')
interaction.plot(mydata$CLASS, mydata$SEX, mydata$VOLUME, fun=mean, col = c('red', 'blue', 'green'), main='Mean Volume per Class', xlab='CLASS', ylab='MEAN VOLUME', trace.label='SEX')
interaction.plot(mydata$CLASS, mydata$SEX, mydata$SHUCK, fun=mean, col = c('red', 'blue', 'green'), main='Mean Shuck per Class', xlab='CLASS', ylab='MEAN SHUCK', trace.label='SEX')
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: For the first plot, it demonstrates that as male abalones age, their ratio of shuck to volume peaks at class A2 and then decreases over time from there. For female abalones, they remain constant until class A2 and then also decrease over time from there. Meanwhile, the infant abalones only decreases over time. The question that gets raised is why are their ratios decreasing? For the second plot, it demonstrates that as abalones mature, they generally grow in volume over time. While the trend for males and infants are very similar, females start off in class A1 with a higher mean volume compared to its counterparts and tend to increase only a little until class A2 while increasing over time. By class A4 and A5, all sex classes' mean volume tend to remain constant. The question that gets raised is why is their growth in mean volume becoming constant in A4 and A5? For the third plot, it demonstrates that as abalones mature, they generally grow in shuck over time. This trend depicted is very similar to the mean volume graph in that while the trend for males and infants are very similar, females again start off in class A1 with a higher mean shuck compared to its counterparts and tend to increase only a little until class A2 while increasing over time. However, at class A4, all sex class' mean shuck decreases until A5. The question that gets raised is why does the mean shuck decrease from A4 to A5?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2, 2))
data <- subset(mydata, mydata$RINGS < 16)
boxplot(data$VOLUME ~ data$RINGS , main = 'Infant Volume | Rings', horizontal = FALSE, col='red', xlab = 'Rings', ylab = 'Volume', subset=data$SEX %in% c('I'), ylim = c(0, 1100))
boxplot(data$VOLUME ~ data$RINGS, main = 'Adult Volume | Rings', horizontal = FALSE, col='blue', xlab = 'Rings', ylab = 'Volume', subset=data$SEX %in% c('F', 'M'), ylim = c(0, 1100))
boxplot(data$WHOLE ~ data$RINGS, main = 'Infant Whole | Rings', horizontal = FALSE, col='red', xlab = 'Rings', ylab = 'Whole Weight', subset=data$SEX %in% c('I'), ylim = c(0, 400))
boxplot(data$WHOLE ~ data$RINGS, main = 'Adult Whole | Rings', horizontal = FALSE, col='blue', xlab = 'Rings', ylab = 'Whole Weight', subset=data$SEX %in% c('F', 'M'), ylim = c(0, 400))
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays suggest that as abalones grow its rings, their volume and whole also increases. However, the growth becomes stagnant after a certain number of rings. Moreover, the difference between infants and adults is that infants have a smaller range in both volume and whole compared to that of adults. This is depicted in the boxplots by the larger boxes as well as the large range of the whiskers for adults whereas in the infant boxplots, there are smaller boxes and shorter whiskers.  ***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: Some plausible statistical reasons that may explain the failure of the original study is the presence of outliers in the data along with the right skewness of the data. These challenges should be addressed before attempting to study this data in order to achieve normality in the data and use approximations of a normal distribution. As witnessed in the prior analysis, physical measurements can only be used for age prediction with younger abalones. Once they reach a certain age, it is difficult to predict their age based solely on their physical attributes since they can be very similar to one another. To be able to use physical measurements, it would be useful to find additional variables that may take environmental factors into account. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Some questions I might ask if I were presented with these information is the size of the sample to make sure that the sample size is representative of the population as a whole as well as the means in collecting these samples to ensure that the means was properly randomized and free of any bias. In the histogram, I would look to identify if there is any skewness to the data and if there are any outliers present to see if the data is normally distributed.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Difficulties analyzing data derived from observational studies include that the data may not consider external or environmental factors that may influence the results of the study. In addition, since these data are purely observational, they may leave room for human error and bias when collecting the data. Outliers and skewness may also be present in the data. Causality cannot be determined since these observations may not be entirely random since the external factors cannot be controlled. The correlations in the data may seem to be identified as causality but it may be due to bias or additional underlying relationships. Such studies depict that it is effective in terms of generalizing relationships between variables but we need a controlled setting to be able to actually draw conclusions. ***
